A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-08-27 manuscript/manuscript.tex
update refs and manuscript

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations using analysis-of-variance. However, differences in means conceal when an experimental effect starts, how long it lasts, and whether its onset is time-locked to other events. Such information is useful not only for interpretation, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).
In this tutorial we focus on a distributional method for analyzing time-to-event data that is known as discrete-time event history analysis (EHA), a.k.a. survival, hazard, duration, failure-time, and transition analysis (Allison, 1982, 2010; Singer \& Willett, 2003). Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of this tutorial is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plan}\NormalTok{(multicore)}
\NormalTok{model\_full\_RE }\OtherTok{\textless{}{-}}
   \FunctionTok{brm}\NormalTok{(}\AttributeTok{data =}\NormalTok{ ptb\_data,}
       \AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link=}\StringTok{"cloglog"}\NormalTok{),}
\NormalTok{       event }\SpecialCharTok{|} \FunctionTok{trials}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ Intercept }\SpecialCharTok{+} 
\NormalTok{                           condition}\SpecialCharTok{*}\NormalTok{period\_9}\SpecialCharTok{*}\NormalTok{trial\_c }\SpecialCharTok{+} 
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                           (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ condition}\SpecialCharTok{*}\NormalTok{period\_9}\SpecialCharTok{*}\NormalTok{trial\_c }\SpecialCharTok{+}
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{|}\NormalTok{ pid),}
       \AttributeTok{prior =}\NormalTok{ priors,}
       \AttributeTok{chains =} \DecValTok{4}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{, }\AttributeTok{iter =} \DecValTok{3000}\NormalTok{, }\AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
       \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.999}\NormalTok{, }\AttributeTok{step\_size =} \FloatTok{0.04}\NormalTok{, }\AttributeTok{max\_treedepth =} \DecValTok{12}\NormalTok{),}
       \AttributeTok{seed =} \DecValTok{12}\NormalTok{, }\AttributeTok{init =} \StringTok{"0"}\NormalTok{,}
       \AttributeTok{file =} \StringTok{"../Tutorial\_2\_Bayesian/models/model\_full\_RE"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

>>>>

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating mean average performance across a series of trials. However, differences in means conceal when an experimental effect starts, how long it lasts, and whether its onset is time-locked to other events. Such information is useful not only for interpretation, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).
In this tutorial we focus on a distributional method for analyzing time-to-event data that is known as discrete-time event history analysis (EHA), a.k.a. survival, hazard, duration, failure-time, and transition analysis (Allison, 1982, 2010; Singer \& Willett, 2003). Across four tutorials that we make publicly available on Github and the Open Science Framework (OSF), we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of this tutorial is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#plan(multicore)}
\CommentTok{\#model\_full\_RE \textless{}{-}}
\CommentTok{\#   brm(data = ptb\_data,}
\CommentTok{\#       family = binomial(link="cloglog"),}
\CommentTok{\#       event | trials(1) \textasciitilde{} 0 + Intercept + }
\CommentTok{\#                           condition*period\_9*trial\_c + }
\CommentTok{\#                           condition*I(period\_9\^{}2) + }
\CommentTok{\#                           condition*I(period\_9\^{}3) +}
\CommentTok{\#                           (1 + condition*period\_9*trial\_c +}
\CommentTok{\#                           condition*I(period\_9\^{}2) +}
\CommentTok{\#                           condition*I(period\_9\^{}3) | pid),}
\CommentTok{\#       prior = priors,}
\CommentTok{\#       chains = 4, cores = 4, iter = 3000, warmup = 1000,}
\CommentTok{\#       control = list(adapt\_delta = 0.999, step\_size = 0.04, max\_treedepth = 12),}
\CommentTok{\#       seed = 12, init = "0",}
\CommentTok{\#       file = "../Tutorial\_2\_Bayesian/models/model\_full\_RE")}
\end{Highlighting}
\end{Shaded}
